MWA tied-array processing III: Microsecond time resolution via a polyphase synthesis filter

### 2020-06-10 referee-report-20200409-response.txt

Dear editor & referee,

We would like to thank both of you for encouraging us to look more deeply into the justification of the synthesis filter that we were using. The result has been not only an improved paper (hopefully!) but also an improved system. We found it interesting to compare the performance of different kinds of filters, and hope that the extra rigour we've inserted into the "Synthesis filters" section will be acceptable to the referee and useful to the readers. On account of the extra work, several sections of the paper have been completely rewritten, as described below in the detailed comments.

The task of getting the improved filter into production was carried out by K. Smith. Because we would not have an in-production system ready by this point if it had not been for his efforts, I would like to request that he be allowed to join the list of authors, as indicated in the manuscript. Please advise us if this is a problem.

@@

Referee: The paper begins with a concise review of PFBs. An overview of synthesis filters used to reverse their operation is given, and the problem is overall well motived. I applaud the authors for laying out the task so clearly.

Referee: The specific choice of synthesis filter, however, seems completely unmotivated and unjustified in any quantitative sense. The authors seem to simply argue "good enough is good enough," but for the central point of the paper I find this woefully insufficient, particularly when well-justified formulations are easily found online and in the community. A quick google reveals https://github.com/jrs65/pfb-inverse as the top result, where a least-squares solution to the inversion is presented. It appears to operate offline, but solves the artefacts described in this work. If the authors require realtime inversion to make it a part of the MWA VCS pipeline, computationally inexpensive solutions exist which vastly reduce the temporal echoes they see. Given that such solutions are in the community already, I would expect at minimum some proof of optimality for some well-chosen metric.

Authors: In hindsight, we completely agree that a paper such as this, if it does not attempt to provide a proper justification for the choice of synthesis filter, cannot possibly offer any value to the radio astronomy community who might be interested in replicating, or improving upon, our method. Accordingly, we investigated the least-squares solution that the referee recommended, and have provided a full, quantitative comparison of multiple synthesis filters, including our original "mirror" filter and a few least-squares solutions of different tap lengths. In so doing, we inevitably found that the least-squares solution performs better than our original filter, and we provide a metric (with accompanying derivation) for comparing them. Armed with this analysis, we implemented (one of) the least squares filters in our software, which is now available. All of this is reflected in the completely written Sections 3 and 4.

@@

Authors: The analysis also revealed that our earlier analysis of the temporal echoes was flawed. Our previous estimates of the magnitude of this effect were greatly exaggerated, and all subsequent test show that the temporal echoes that we see are consistent with the $\lesssim -25$ dB "ringing" shown in the bottom panel of (the new) Fig. 4, which is a general feature of all imperfect back-to-back systems. Accordingly, we no longer had any reason to suppose that any of the features we saw in our pulsar observations were due to this effect, and have removed the relevant sentences from that section.

Referee: I was glad to see the impulse response to characterize performance, but the comparison to an ifft seems silly except to point out that one should not expect iffts to invert pfbs.

Authors: Yes, we agree. The comparison with the ifft is not useful, and we have removed those paragraphs. However, the referee will note that the new version no longer contains an impulse response test. The justification for this decision we give in the paper itself (the second paragraph of Section 4), which amounts to an argument that the impulse response is not the best way to characterise the performance of our particular system, where significant non-linearities are introduced by rounding effects. We therefore focus our efforts on the empirical test done on actual data (Section 4), being careful to quantify its performance in terms of the same metric introduced earlier (Section 3.1).

@@

Referee: The pulsar observations constitute a nice proof-of-concept, and serve as demonstrations of the chosen filter in the field as the authors do not pursue them to novel results. I do not understand the comparison of incoherently- to coherently-dedispersed profiles, it seems very much apples-to-oranges and not directly relevant to the variable time/freq division of the data. The time-resolved sub-pulse microstructure of B0950+08, however, is an impressive demonstration, well worth including in a revised paper.

Authors: The relevance of the comparison between coherent and incoherent de-dispersion lies in the fact having a high time resolution is useless if the resolution is limited by dispersive smearing across a fine channel, which can easily become the dominant factor for MSPs de-dispersed incoherently. In the original (non-high-time resolution) mode, coherent de-dispersion was not an option, and we wanted to showcase the difference. However, the point is well taken: this has more to do with the limitations of the downstream analysis, and not so much with the synthesis system itself. In order not to belabour that point, we have removed the incoherently de-dispersed version of the J2241−5236 plot (Fig 7); however, we have left both J0437-4715 plots in the paper (Fig 8), in order to help the reader identify the "notches" discussed in the text.

Referee: Lastly, I would be remiss if I didn't express some concern that the PFB implemented on the FPGAs cannot be exactly reproduced on general-purpose compute hardware! The difference in hardware can hardly matter for a digital operation, and myriad libraries exist to implement arbitrary bit-depth arithmetic in CPUs. It's clear from the properties presented in appendix A that a number of awkward features are already reproduced, and I'd strongly encourage the authors to find an FFT implementation (or implement one themselves) which perfectly mimics that on the FPGA.

Authors: Finding and running the correct version of the fixed-point FFT implemented on the FPGAs proved difficult: that version is no longer supported, and our early attempts to replicate it were not successful... although in honesty we cannot be 100% sure of the reason. Neverthess, this point may now be moot. If the referee accepts our arguments that the impulse response test does not adequately characterise the system, then there is no longer any need to reproduce the forward PFB. Instead, the empirical test we offer instead uses the actual implemented system, meaning that the reported performance includes any peculiarites introduced by the FPGA (related to the FFT or otherwise). Obviously it's less than satisfactory not to have a complete account of what the forward PFB is doing, but our investigations indicate that the rounding errors described in Section 4 and in the Appendix account for virtually the entire error budget (beyond that introduced by the choice of synthesis filter itself). Any deeper investigation would only be warranted if there was some indication that such effects could be mitigated, which we don't find here.

### 2020-07-07 ipfb.tex
Accepted version (without the boldface change markups)

\newcommand{\changes}[1]{{\color{red}#1}}

@@

\author[McSweeney et al.]{S. J. McSweeney$^1$, S. M. Ord$^2$, D. Kaur$^1$, N. D. R. Bhat$^1$, B. W. Meyers$^{1,2,3}$, S. E. Tremblay, J. Jones$^1$, B. Crosse$^1$, K. Smith$^1$
\affil{$^1$International Centre for Radio Astronomy Research (ICRAR), Curtin University, 1 Turner Avenue, Technology Park, Bentley, 6102, W.A., Australia}
\affil{$^2$CSIRO Astronomy and Space Science, PO Box 76, Epping, NSW 1710, Australia}
\affil{$^3$Department of Physics and Astronomy, University of British Columbia, 6224 Agricultural Road, Vancouver, BC V6T 1Z1, Canada}
}

@@

Fig. \ref{fig:snr} suggests that one can achieve arbitrarily good performance by choosing a sufficiently long filter.
However, the better performance of the longer filters is offset by the increased computational resources and/or time required to apply them, which may exceed the system's design specifications.
Note that the mirror filter, being the same size as the 12-tap least squares filter, offers no advantage in terms of minimising the reconstruction error.
On the other hand, the way that the signal power is distributed across multiple taps is markedly different for the two filter types, as shown in the bottom panel of Fig. \ref{fig:snr}.
The mirror filter performs slightly worse in nearby taps, but drops off more quickly further out.
For applications where the samples are signal-dominated, and minimising temporal imaging is more important than measuring the S/N of the signal precisely, the mirror filter may therefore offer some advantage over the least squares solution.
However, observations of pulsars only rarely fall into this regime, even for observations of single pulses, where typically many samples must be averaged before the signal becomes more significant than the noise.
A more detailed analysis of these effects are therefore outside the scope of the present work.

@@

\psrslowB{} is a bright, long-period ($P = 0.253\,$s) pulsar known to exhibit microstructure \citep{Popov2002,Kuzmin2003}.
A total of eighty seconds (315 pulses) of data ($128 \times 10\,$kHz fine channels) were recorded, beamformed, and subjected to the polyphase synthesis filter.
Across the MWA bandwidth, we used PSRCHIVE's RMFIT to measure an approximate rotation measure (RM) of $1.43 \pm 0.15\,$rad m$^{-2}$, which is consistent with previous measurements of this pulsar's RM \citep[e.g.][]{Noutsos2015}.
The mean profile formed from the one coarse channel is shown in Fig. \ref{fig:0950_80_secs}.
Several individual pulses were sufficiently bright to see substructure.
One such pulse is showcased in Fig. \ref{fig:0950_single_pulse}, where all 24 coarse channels were integrated to maximise the S/N.
To highlight the pulse's substructure, we show it at three different time resolutions (corresponding to the chosen bin width of each plot).

>>>>

@@

\author[McSweeney et al.]{S. J. McSweeney$^1$, S. M. Ord$^2$, D. Kaur$^1$, N. D. R. Bhat$^1$, B. W. Meyers$^{1,2,3}$, S. E. Tremblay, J. Jones$^1$, B. Crosse$^1$, K. R. Smith$^1$
\affil{$^1$International Centre for Radio Astronomy Research (ICRAR), Curtin University, 1 Turner Avenue, Technology Park, Bentley, 6102, W.A., Australia}
\affil{$^2$CSIRO Astronomy and Space Science, PO Box 76, Epping, NSW 1710, Australia}
\affil{$^3$Department of Physics and Astronomy, University of British Columbia, 6224 Agricultural Road, Vancouver, BC V6T 1Z1, Canada}
}

@@

Fig. \ref{fig:snr} suggests that one can achieve arbitrarily good performance by choosing a sufficiently long filter.
However, the better performance of the longer filters is offset by the increased computational resources and/or time required to apply them, which may exceed the system's design specifications.
Note that the mirror filter, being the same size as the 12-tap least squares filter, offers no advantage in terms of minimising the reconstruction error.
On the other hand, the way that the signal power is distributed across multiple taps is markedly different for the two filter types, as shown in the bottom panel of Fig. \ref{fig:snr}.
The mirror filter performs slightly worse in nearby taps, but drops off more quickly further out.
For applications where the samples are signal-dominated, and minimising temporal imaging is more important than measuring the S/N of the signal precisely, the mirror filter may therefore offer some advantage over the least squares solution.
However, observations of pulsars only rarely fall into this regime, even for observations of single pulses, where typically many samples must be averaged before the signal becomes more significant than the noise.
A more detailed analysis of these effects is therefore outside the scope of the present work.

@@

\psrslowB{} is a bright, long-period ($P = 0.253\,$s) pulsar known to exhibit microstructure \citep{Popov2002,Kuzmin2003}.
A total of eighty seconds (315 pulses) of data ($128 \times 10\,$kHz fine channels) were recorded, beamformed, and subjected to the polyphase synthesis filter.
Across the MWA bandwidth, we used the RM synthesis technique\footnote{\url{https://github.com/gheald/RMtoolkit}} \citep{Brentjens2005} to measure a rotation measure (RM) of $1.43 \pm 0.15\,$rad m$^{-2}$, which is consistent with previous measurements of this pulsar's RM \citep[e.g.][]{Noutsos2015}.
The mean profile formed from the one coarse channel is shown in Fig. \ref{fig:0950_80_secs}.
Several individual pulses were sufficiently bright to see substructure.
One such pulse is showcased in Fig. \ref{fig:0950_single_pulse}, where all 24 coarse channels were integrated to maximise the S/N.
To highlight the pulse's substructure, we show it at three different time resolutions (corresponding to the chosen bin width of each plot).
